Simple repression draft

### 2020-09-11 doc/resubmission/appendix_01_nonbursty.tex
proposed changes to beyond means section

>>>>

\mrm{THIS NEEDS TO BE FIXED}
\subsubsection{Noise in the Two-State Promoter, RNAP Bound or Unbound.}

Our second model of constitutive transcription posits an architecture in which
the promoter is either empty or bound by RNAP ~\cite{Phillips2015a,
Phillips2019}. Here,  as shown in model 2 of Figure~\ref{fig2:constit_cartoons}(A),
transcription initiation results in a state transition from the bound to the
unbound state, reflecting the microscopic reality that an RNAP that has begun to
elongate a transcript is no longer available at the start site to begin another.
As shown in Appendix~\ref{sec:non_bursty}, the Fano factor in this model is
given by
\begin{align}
    \nu = 1 -
        \frac{r\rate{k}{P}{+}}
            {\left(\rate{k}{P}{+} + \rate{k}{P}{-} + r\right)
             \left(\gamma + \rate{k}{P}{+} + \rate{k}{P}{-} + r\right)}.
\label{eq:model2_fano}
\end{align}
The problem with this picture is that the Fano factor is always $<1$. To make
 contact with the experimental reality of $\nu>1$ as shown in
 Figure~\ref{fig2:constit_cartoons}(B), clearly some corrections will be needed.
 While this model adds an appealing element of microscopic reality, we are
 forced to reject it as the additional complexity is unable to capture the
 phenomenology of interest. Obviously the promoter state does in fact proceed
 through cycles of RNAP binding, initiating, and elongating, but it seems that
 the super-Poissonian noise in mRNA copy number we want to model must be
 governed by other features of the system.

\subsubsection{Noise in the Three-State Promoter, Multistep Transcription
Initiation and Escape.}

How might we remedy the deficits of model 2? It is known~\cite{DeHaseth1998}
that once RNAP initially binds the promoter region, a multitude of distinct
steps occur sequentially before RNAP finally escapes into the elongation phase.
Perhaps adding some of this mechanistic detail as shown in model 3 of
Figure~\ref{fig2:constit_cartoons}(A) might rescue the previous model. The next
simplest refinement of that model could consider open complex formation and
promoter escape as separate steps rather than as a single effective step. In
other words, we construct model 3 by adding a single extra state to model 2, and
we will label the two RNAP-bound states as the closed and open complexes,
despite the true biochemical details certainly being more complex. For example,
earlier work extended this model by adding an additional repressor bound state
and did not explicitly consider the limit with no repressor that we analyze
here~\cite{Mitarai2015}. Again, our goal here is not a complete accounting of
all the relevant biochemical detail; this is an exploratory search for the
important features that a model needs to include to square with the known
experimental reality of constitutive expression.

Unfortunately, as hinted at in earlier work~\cite{Mitarai2015}, this model too
has Fano factor $\nu<1$. We again leave the algebraic details for
Appendix~\ref{sec:non_bursty} and merely state the result that
\begin{align}
\nu = 1 - \frac{r k_P^+ k_O}{\mathcal{Z}}
\frac{k_P^+ + k_P^- + k_O + r + \gamma}
    {\mathcal{Z} + \gamma(k_P^+ + k_P^- + k_O + r) + \gamma^2},
\label{eq:model3_fano}
\end{align}
where we defined $\mathcal{Z} = r(k_O + k_P^-) + k_P^+(k_O + r)$ for notational
tidiness. This is necessarily less than 1 for arbitrary rate constants.

In fact, we suspect \textit{any} model in which transcription proceeds through a
multistep cycle must necessarily have $\nu<1$. The intuitive argument compares
the waiting time distribution to traverse the cycle with the waiting time for a
Poisson promoter (model 1) with the same mean time. The latter is simply an
exponential distribution. The former is a convolution of multiple exponentials,
and intuitively the waiting time distribution for a multistep process
should be more peaked with a smaller fractional width than a single
exponential with the same mean.
A less disperse waiting time distribution means transcription initations are
more uniformly distributed in time relative to a Poisson process. Hence the
distribution of mRNA over a population of cells should be less variable compared
to Poisson, giving $\nu<1$.
(In Appendix~\ref{sec:non_bursty} we present a more precise version
of the intuitive arguments in this paragraph.)
Regardless of the merits of this model in describing the noise properties of
constitutive transcription initiation, it ultimately fails the simplest
quantitative feature of the data, namely that the Fano factor $> 1$ and hence
we must discard this mechanistic picture and search elsewhere.
\mrm{ENDS HERE}

### 2020-09-11 doc/resubmission/section_02_means.tex
proposed changes to beyond means section

\subsubsection{Two- and Three-state Thermodynamic Models}
We begin our analysis with models (1) and (2) in
Figure~\ref{fig1:means_cartoons}(B). In each of these models the promoter
is idealized as existing in a set of discrete states; the difference being
whether or not the RNAP bound state is included or not. Gene expression is then
assumed to be proportional to the probability of the promoter being in either
the empty state (model (1)) or the RNAP-bound state (model (2)). We direct the
reader to Appendix~\ref{sec:non_bursty} for details on the derivation of the
fold-change. For our purposes here, it suffices to state that the functional
form of the fold-change for model (1) is
\begin{equation}
\text{fold-change}
= \left(1 + \frac{R}{N_{NS}} e^{-\beta\Delta\varepsilon_R}\right)^{-1},
\end{equation}
where $R$ is the number of repressors per cell, $N_{NS}$ is the number of
non-specific binding sites where the repressor can bind, $\Delta\varepsilon_R$
is the repressor-operator binding energy, and $\beta \equiv (k_BT)^{-1}$. This
equation matches the form of the master curve in
Figure~\ref{fig1:means_cartoons}(D) with $\rho=1$ and $\Delta F_R =
\beta\Delta\varepsilon_r - \log (R / N_{NS})$. For model (2) we have a similar
situation. The fold-change takes the form
\begin{align}
\text{fold-change}
&= \left(
1 + \frac{\frac{R}{N_{NS}} e^{-\beta\Delta\varepsilon_R}}
        {1 + \frac{P}{N_{NS}} e^{-\beta\Delta\varepsilon_P}}
\right)^{-1}
\\
&= (1 + \exp(-\Delta F_R + \log\rho))^{-1},
\end{align}
where $P$ is the number of RNAP per cell, and $\Delta\varepsilon_P$ is the
RNAP-promoter binding energy. For this model we have $\Delta F_R =
\beta\Delta\varepsilon_R - \log(R/N_{NS})$ and $\rho = 1 +
\frac{P}{N_{NS}}\mathrm{e}^{-\beta\Delta\varepsilon_P}$. Thus far, we see that
the two thermodynamic models, despite making different coarse-graining
commitments, result in the same functional form for the fold-change in mean gene
expression.  We now explore how kinetic models fare when faced with computing
the same observable.

>>>>

\subsubsection{Two- and Three-state Thermodynamic Models}
We begin our analysis with models 1 and 2 in
Figure~\ref{fig1:means_cartoons}(B). In each of these models the promoter
is idealized as existing in a set of discrete states; the difference being
whether or not the RNAP bound state is included or not. Gene expression is then
assumed to be proportional to the probability of the promoter being in either
the empty state (model 1) or the RNAP-bound state (model (2)). We direct the
reader to Appendix~\ref{sec:non_bursty} for details on the derivation of the
fold-change. For our purposes here, it suffices to state that the functional
form of the fold-change for model 1 is
\begin{equation}
\text{fold-change}
= \left(1 + \frac{R}{N_{NS}} e^{-\beta\Delta\varepsilon_R}\right)^{-1},
\end{equation}
where $R$ is the number of repressors per cell, $N_{NS}$ is the number of
non-specific binding sites where the repressor can bind, $\Delta\varepsilon_R$
is the repressor-operator binding energy, and $\beta \equiv (k_BT)^{-1}$. This
equation matches the form of the master curve in
Figure~\ref{fig1:means_cartoons}(D) with $\rho=1$ and $\Delta F_R =
\beta\Delta\varepsilon_r - \log (R / N_{NS})$. For model 2 we have a similar
situation. The fold-change takes the form
\begin{align}
\text{fold-change}
&= \left(
1 + \frac{\frac{R}{N_{NS}} e^{-\beta\Delta\varepsilon_R}}
        {1 + \frac{P}{N_{NS}} e^{-\beta\Delta\varepsilon_P}}
\right)^{-1}
\\
&= (1 + \exp(-\Delta F_R + \log\rho))^{-1},
\end{align}
where $P$ is the number of RNAP per cell, and $\Delta\varepsilon_P$ is the
RNAP-promoter binding energy. For this model we have $\Delta F_R =
\beta\Delta\varepsilon_R - \log(R/N_{NS})$ and $\rho = 1 +
\frac{P}{N_{NS}}\mathrm{e}^{-\beta\Delta\varepsilon_P}$. Thus far, we see that
the two thermodynamic models, despite making different coarse-graining
commitments, result in the same functional form for the fold-change in mean gene
expression.  We now explore how kinetic models fare when faced with computing
the same observable.

### 2020-09-11 doc/resubmission/section_03_beyond_means.tex
implement Rob's comments
proposed changes to beyond means section

\begin{figure}%[h!]
\centering
\includegraphics[width=\textwidth]{../figures/main/fig02.pdf}
\caption{\textbf{Comparison of different models for noise in the constitutive promoter.}
(A) The left column depicts various plausible models for the dynamics of
constitutive promoters. In model (1), transcripts are produced in a Poisson
process~\cite{Sanchez2013, Jones2014}. Model (2) features explicit modeling of
RNAP binding/unbinding kinetics~\cite{Phillips2015a}. Model (3) is a more
detailed generalization of model (2), treating transcription initiation as a
multi-step process proceeding through closed and open
complexes~\cite{Mitarai2015}. Model (4) is somewhat analogous to (2) except with
the precise nature of active and inactive states left
ambiguous~\cite{Peccoud1995, Shahrezaei2008, Razo-Mejia2020}. Finally, model (5)
can be viewed as a certain limit of model (4) in which transcripts are produced
in bursts, and initiation of bursts is a Poisson process. 
% \mmnote{Haven't found model 5 formulated quite like this in the literature,
% usually burstiness is handled like model 4, but I should probably search a bit
% more.} 
The right column shows the Fano factor $\nu$ (variance/mean) for each model.
Note especially the crucial diagnostic: (2) and (3) have $\nu$ strictly below 1,
while only for (4) and (5) can $\nu$ exceed 1. Models with Fano factors $\le 1$
cannot produce the single-cell data observed in part (B) without introducing
additional assumptions and model complexity. (B) Data from \cite{Jones2014}.
Mean mRNA count vs. Fano factor (variance/mean) for different promoters as
determined with single-molecule mRNA Fluorescence \textit{in situ}
Hybridization. The colorbar indicates the predicted binding affinity of RNAP to
the promoter sequence as determined in \cite{Brewster2012}. Numbers serve for
cross comparison with data presented in Figure 3.}
\label{fig2:constit_cartoons}
\end{figure}

\subsubsection{Noise in the Poisson Promoter Model}

The simplest model of constitutive expression that we can imagine is shown as
model 1 in Figure~\ref{fig2:constit_cartoons}(A) and assumes that transcripts are
produced as a Poisson process from a single promoter state. This is the picture
from Jones et.\ al.~\cite{Jones2014} that was used to interpret a systematic
study of gene expression noise over a series of promoters designed to have
different strengths. This model insists that the ``true'' steady-state mRNA
distribution is Poisson, implying the Fano factor $\nu$ must be 1.
In~\cite{Jones2014}, the authors carefully attribute measured deviations from
Fano = 1 to intensity variability in fluorescence measurements, gene copy number
variation, and copy number fluctuations of the transcription machinery, e.g.,
RNAP itself. In this picture, the master equation makes no appearance, and all
the corrections to Poisson behavior are derived as additive corrections to the
Fano factor. For disproving the ``universal noise curve'' from So et.\
al.~\cite{So2011}, this picture was excellent. It is appealing in its simplicity,
with only two parameters, the initiation rate $r$ and degradation rate $\gamma$.
Since $\gamma$ is independently known from other experiments, and the mean mRNA
copy number is $r/\gamma$, $r$ is easily inferred from data. In other words, the
model is not excessively complex for the data at hand. But for many interesting
questions, for instance in the recent work~\cite{Razo-Mejia2020}, knowledge of
means and variances alone is insufficient, and a description of the full
distribution of molecular counts is necessary. For this we need a (slightly)
more complex model than model 1 that would allow us to incorporate the
non-Poissonian features of constitutive promoters directly into a master
equation formulation.

\subsubsection{Noise in the Two-State Promoter, RNAP Bound or Unbound.}

Our second model of constitutive transcription posits an architecture in which
the promoter is either empty or bound by RNAP ~\cite{Phillips2015a,
Phillips2019}. Here,  as shown in model 2 of Figure~\ref{fig2:constit_cartoons}(A),
transcription initiation results in a state transition from the bound to the
unbound state, reflecting the microscopic reality that an RNAP that has begun to
elongate a transcript is no longer available at the start site to begin another.
As shown in Appendix~\ref{sec:non_bursty}, the Fano factor in this model is
given by
\begin{align}
    \nu = 1 -
        \frac{r\rate{k}{P}{+}}
            {\left(\rate{k}{P}{+} + \rate{k}{P}{-} + r\right)
             \left(\gamma + \rate{k}{P}{+} + \rate{k}{P}{-} + r\right)}.
\label{eq:model2_fano}
\end{align}
The problem with this picture is that the Fano factor is always $<1$. To make
 contact with the experimental reality of $\nu>1$ as shown in
 Figure~\ref{fig2:constit_cartoons}(B), clearly some corrections will be needed.
 While this model adds an appealing element of microscopic reality, we are
 forced to reject it as the additional complexity is unable to capture the
 phenomenology of interest. Obviously the promoter state does in fact proceed
 through cycles of RNAP binding, initiating, and elongating, but it seems that
 the super-Poissonian noise in mRNA copy number we want to model must be
 governed by other features of the system.

\subsubsection{Noise in the Three-State Promoter, Multistep Transcription
Initiation and Escape.}

How might we remedy the deficits of model 2? It is known~\cite{DeHaseth1998}
that once RNAP initially binds the promoter region, a multitude of distinct
steps occur sequentially before RNAP finally escapes into the elongation phase.
Perhaps adding some of this mechanistic detail as shown in model 3 of
Figure~\ref{fig2:constit_cartoons}(A) might rescue the previous model. The next
simplest refinement of that model could consider open complex formation and
promoter escape as separate steps rather than as a single effective step. In
other words, we construct model 3 by adding a single extra state to model 2, and
we will label the two RNAP-bound states as the closed and open complexes,
despite the true biochemical details certainly being more complex. For example,
earlier work extended this model by adding an additional repressor bound state
and did not explicitly consider the limit with no repressor that we analyze
here~\cite{Mitarai2015}. Again, our goal here is not a complete accounting of
all the relevant biochemical detail; this is an exploratory search for the
important features that a model needs to include to square with the known
experimental reality of constitutive expression.

Unfortunately, as hinted at in earlier work~\cite{Mitarai2015}, this model too
has Fano factor $\nu<1$. We again leave the algebraic details for
Appendix~\ref{sec:non_bursty} and merely state the result that
\begin{align}
\nu = 1 - \frac{r k_P^+ k_O}{\mathcal{Z}}
\frac{k_P^+ + k_P^- + k_O + r + \gamma}
    {\mathcal{Z} + \gamma(k_P^+ + k_P^- + k_O + r) + \gamma^2},
\label{eq:model3_fano}
\end{align}
where we defined $\mathcal{Z} = r(k_O + k_P^-) + k_P^+(k_O + r)$ for notational
tidiness. This is necessarily less than 1 for arbitrary rate constants.

In fact, we suspect \textit{any} model in which transcription proceeds through a
multistep cycle must necessarily have $\nu<1$. The intuitive argument compares
the waiting time distribution to traverse the cycle with the waiting time for a
Poisson promoter (model 1) with the same mean time. The latter is simply an
exponential distribution. The former is a convolution of multiple exponentials,
and intuitively the waiting time distribution for a multistep process
should be more peaked with a smaller fractional width than a single
exponential with the same mean.
A less disperse waiting time distribution means transcription initations are
more uniformly distributed in time relative to a Poisson process. Hence the
distribution of mRNA over a population of cells should be less variable compared
to Poisson, giving $\nu<1$.
(In Appendix~\ref{sec:non_bursty} we present a more precise version
of the intuitive arguments in this paragraph.)
Regardless of the merits of this model in describing the noise properties of
constitutive transcription initiation, it ultimately fails the simplest
quantitative feature of the data, namely that the Fano factor $> 1$ and hence
we must discard this mechanistic picture and search elsewhere.

@@

One interpretation~\cite{Chong2014, Sevier2016} for the active and inactive
states is that they represent the promoter's supercoiling state: transitions to
the inactive state are caused by accumulation of positive supercoiling, which
inhibits transcription, and transitions back to ``active'' are caused by gyrase
or other topoisomerases relieving the supercoiling. This is an interesting
possibility because it would mean the timescale for promoter state transitions
is driven by topoisomerase kinetics, not by RNAP kinetics. From in vitro
measurements, the former are suggested to be of order minutes~\cite{Chong2014}.
Contrast this with model 2, where the state transitions are assumed to be
governed by RNAP, which, assuming a copy number per cell of order $10^3$, has a
diffusion-limited association rate $k_{on} \sim 10^2~\text{s}^{-1}$ to a target
promoter. Combined with known $K_d$'s of order $\mu$M, this gives an RNAP
dissociation rate $k_{off}$ of order $10^2$~s$^{-1}$. As we will show below,
however, there are some lingering puzzles with interpreting this supercoiling
hypothesis, so we leave it as a speculation and refrain from assigning definite
physical meaning to the two states in this model.

@@

% \marginpar{\it RP: best description I have ever seen of this.  Good job}
To develop some intuition, consider model 4 in the limit $k^+ \ll k^- \lesssim
r$, which is roughly satisfied by the parameters inferred
in~\cite{Razo-Mejia2020}. In this limit, the system spends the majority of its
time in the inactive state, occasionally becoming active and making a burst of
transcripts. This should call to mind the well-known phenomenon of
transcriptional bursting, as reported in,
e.g.,~\cite{Golding2005,Chong2014,Sevier2016}.
%\mmnote{should probably add some more cites here}.
Let us make this correspondence more precise. The mean dwell
time in the active state is $1/k^-$. While in this state, transcripts are
produced at a rate $r$ per unit time. So on average, $r/k^-$ transcripts are
produced before the system switches to the inactive state. Once in the inactive
state, the system dwells there for an average time $1/k^+$ before returning to
the active state and repeating the process. $r/k^-$ resembles an average burst
size, and $1/k^+$ resembles the time interval between burst events. More
precisely, $1/k^+$ is the mean time between the end of one burst and the start
of the next, whereas $1/k^+ + 1/k^-$ would be the mean interval between the
start of two successive burst events, but in the limit $k^+ \ll k^-$, $1/k^+ +
1/k^- \approx 1/k^+$. Note that this limit ensures that the waiting time between
bursts is approximately exponentially distributed, with mean set by the only
timescale left in the problem, $1/k^+$.
% \footnote{If instead it were the case that $k^+ \sim k^-$, then the waiting time
% $1/k^+ + 1/k^-$ would have a peaked distribution, but this does not appear to be
% the case for any of datasets from~\cite{Jones2014}.}

Let us now verify this intuition with a precise derivation to check that $r/k^-$
is in fact the mean burst size and to obtain the full burst size distribution.
Consider first a constant, known dwell time $T$ in the active state. Transcripts
are produced at a rate $r$ per unit time, so the number of transcripts $n$
produced during $T$ is Poisson distributed with mean $rT$, i.e.,
\begin{equation}
    P(n\mid T) = \frac{(rT)^n}{n!} \exp(-rT).
\end{equation}
Since the dwell time $T$ is unobservable, we actually want $P(n)$, the dwell
time distribution with no conditioning on $T$. Basic rules of probability theory
tell us we can write $P(n)$ in terms of $P(n\mid T)$ as
\begin{equation}
    P(n) =\int_0^\infty P(n\mid T) P(T) dT.
\end{equation}
But we know the dwell time distribution $P(T)$, which is exponentially
distributed according to
\begin{equation}
    P(T) = k^- \exp(-T k^-),
\end{equation}
so $P(n)$ can be written as
\begin{equation}
    P(n) = k^- \frac{r^n}{n!}
            \int_0^\infty T^n\exp[-(r + k^-)T]\,dT.
\end{equation}
A standard integral table shows $\int_0^\infty x^n e^{-ax}\,dx = n!/a^{n+1}$, so
\begin{equation}
    P(n) = \frac{k^- r^n}{(k^- + r)^{n+1}}
        = \frac{k^-}{k^- + r}
            \left(\frac{r}{k^- + r}\right)^n
        = \frac{k^-}{k^- + r}
            \left(1 - \frac{k^-}{k^- + r}\right)^n,
\end{equation}
which is exactly the geometric distribution with standard parameter
$\theta\equiv k^-/(k^- + r)$ and domain $n \in \{0, 1, 2, \dots\}$.
The mean of the geometric distribution, with this convention, is
\begin{align}
\langle n\rangle = \frac{1 - \theta}{\theta}
        = \left(1 - \frac{k^-}{k^- + r}\right)
                    \frac{k^- + r}{k^-}
        = \frac{r}{k^-},
\end{align}
exactly as we guessed intuitively above.

@@

Consider the time delay between the first and final RNAPs in a burst initiating
transcription (\textit{not} the time to complete transcripts, which potentially
could be much longer.) If this timescale is short compared to the typical search
timescale of transcription factors, then all is well. The estimates from
deHaseth et.\ al.~\cite{DeHaseth1998} put RNAP's diffusion-limited on rate
around $\sim\text{few}\times10^{-2}~\text{nM}^{-1}~\text{s}^{-1}$ and polymerase
loading as high as $1~\text{s}^{-1}$. Then for reasonable burst sizes of $<10$,
it is reasonable to guess that bursts might finish initiating on a timescale of
tens of seconds or less (with another 30-60 sec to finish elongation, but that
does not matter here). A transcription factor with typical copy number of order
10 (or less) would have a diffusion-limited association rate of order
$(10~\text{sec})^{-1}$ \cite{Hammar2014}. Higher copy number TFs tend to have
many binding sites over the genome, which should serve to pull them out of
circulation and keep their effective association rates from rising too large.
Therefore, there is \textit{perhaps} a timescale separation possible between
transcription factor association rates and burst durations, but this assumption
could very well break down, so we will have to keep it in mind when we infer
repressor rates from the Jones et.\ al.\ single-cell mRNA counts data
later~\cite{Jones2014}.

>>>>

\begin{figure}%[h!]
\centering
\includegraphics[width=\textwidth]{../../figures/main/fig02.pdf}
\caption{\textbf{Comparison of different models for noise in the constitutive promoter.}
(A) The left column depicts various plausible models for the dynamics of
constitutive promoters. In model (1), transcripts are produced in a Poisson
process~\cite{Sanchez2013, Jones2014}. Model (2) features explicit modeling of
RNAP binding/unbinding kinetics~\cite{Phillips2015a}. Model (3) is a more
detailed generalization of model (2), treating transcription initiation as a
multi-step process proceeding through closed and open
complexes~\cite{Mitarai2015}. Model (4) is somewhat analogous to (2) except with
the precise nature of active and inactive states left
ambiguous~\cite{Peccoud1995, Shahrezaei2008, Razo-Mejia2020}. Finally, model (5)
can be viewed as a certain limit of model (4) in which transcripts are produced
in bursts, and initiation of bursts is a Poisson process. 
% \mmnote{Haven't found model 5 formulated quite like this in the literature,
% usually burstiness is handled like model 4, but I should probably search a bit
% more.} 
The right column shows the Fano factor $\nu$ (variance/mean) for each model.
Note especially the crucial diagnostic: (2) and (3) have $\nu$ strictly below 1,
while only for (4) and (5) can $\nu$ exceed 1. Models with Fano factors $\le 1$
cannot produce the single-cell data observed in part (B) without introducing
additional assumptions and model complexity. (B) Data from \cite{Jones2014}.
Mean mRNA count vs. Fano factor (variance/mean) for different promoters as
determined with single-molecule mRNA Fluorescence \textit{in situ}
Hybridization. The colorbar indicates the predicted binding affinity of RNAP to
the promoter sequence as determined in \cite{Brewster2012}. Numbers serve for
cross comparison with data presented in Figure 3.}
\label{fig2:constit_cartoons}
\end{figure}

\subsubsection{Poisson Noise promoter}

The simplest model of constitutive expression that we can imagine is shown as
model 1 in Figure~\ref{fig2:constit_cartoons}(A) and assumes that transcripts
are produced as a Poisson process from a single promoter state. This is the
picture from Jones et.\ al.~\cite{Jones2014} that was used to interpret a
systematic study of gene expression noise over a series of promoters designed to
have different strengths. This model insists that the ``true'' steady-state mRNA
distribution is Poisson, implying the Fano factor $\nu$ must be 1.
In~\cite{Jones2014}, the authors carefully attribute measured deviations from
Fano = 1 to intensity variability in fluorescence measurements, gene copy number
variation, and copy number fluctuations of the transcription machinery, e.g.,
RNAP itself. In this picture, the master equation makes no appearance, and all
the corrections to Poisson behavior are derived as additive corrections to the
Fano factor. For \mrm{challenging} the ``universal noise curve'' from So et.\
al.~\cite{So2011}, this picture was excellent. It is appealing in its
simplicity, with only two parameters, the initiation rate $r$ and degradation
rate $\gamma$. Since $\gamma$ is independently known from other experiments, and
the mean mRNA copy number is $r/\gamma$, $r$ is easily inferred from data. In
other words, the model is not excessively complex for the data at hand. But for
many interesting questions, for instance in the recent
work~\cite{Razo-Mejia2020}, knowledge of means and variances alone is
insufficient, and a description of the full distribution of molecular counts is
necessary. For this we need a (slightly) more complex model than model 1 that
would allow us to incorporate the non-Poissonian features of constitutive
promoters directly into a master equation formulation.

\mrm{I added the word challenging instead of disproving because of the 
following comment from reviewer 1. But maybe more needs to be discussed about 
this?}

\mrm{
    “disproving the universal noise curve from So et. al. [59]” (page 18). How was the universal noise curve disproved? Universal noise/mean relations, similar to those reported in ref \#59, were repeatedly observed, in both bacteria and eukaryotes (see, e.g., PMID: 24311680, PMID: 25858977). Moreover, later reports showed that the trend is observed even when accounting for gene dosage effects (PMID: 26965629). Thus, the statement above requires clarification.
}

\subsubsection{Sub-Poissoninan noise promoters with RNAP escape}
A natural extension of the one-state promoter studied in the previous section is
to explicitly include an empty promoter state. This state allows for single RNAP
to bind and unbind from the promoter with rates $k_P^+$ and $k_P^-$,
respectively, before engaging in a transcriptional event. Once the RNAP
irreversibly escapes from the promoter to transcribe the mRNA, the promoter goes
back to this empty state. Models 2 and 3 from
Figure~\ref{fig2:constit_cartoons}(A) show two different versions of this
process. In model 2 the polymerase can directly engage in a transcription event
with rate $r$ after binding to the promoter, while model 3 includes an
irreversible step with rate $k_O$ that coarse-grains the known multi-step
process involved in transcriptional initiation~\cite{DeHaseth1998}.

Although physically intuitive given our understanding of how a single transcript
is produced, we are forced to discard both of these models on the basis that the
Fano factor for both of them is strictly $< 1$ as shown in
Figure~\ref{fig2:constit_cartoons}(A) (See Appendix~\ref{sec:non_bursty} for
detailed derivations). This is in disagreement with the experimental data
from~\cite{Jones2014}, reproduced in Figure~\ref{fig2:constit_cartoons}(B). All
of these unregulated promoters --which only differ on the promoter sequence--
show strictly super-Poissonian noise (Fano factor $> 1$). In fact, we suspect
\textit{any} model in which transcription proceeds through a multistep cycle
must necessarily have $\nu<1$. The intuitive argument compares the waiting time
distribution to traverse the cycle with the waiting time for a Poisson promoter
(model 1) with the same mean time. The latter is simply an exponential
distribution. The former is a convolution of multiple exponentials, and
intuitively the waiting time distribution for a multistep process should be more
peaked with a smaller fractional width than a single exponential with the same
mean. A less disperse waiting time distribution means transcription initations
are more uniformly distributed in time relative to a Poisson process. Hence the
distribution of mRNA over a population of cells should be less variable compared
to Poisson, giving $\nu<1$. (In Appendix~\ref{sec:non_bursty} we present a more
precise version of the intuitive arguments in this paragraph).

@@

One interpretation~\cite{Chong2014, Sevier2016} for the active and inactive
states is that they represent the promoter's supercoiling state: transitions to
the inactive state are caused by accumulation of positive supercoiling, which
inhibits transcription, and transitions back to ``active'' are caused by gyrase
or other topoisomerases relieving the supercoiling. This is an interesting
possibility because it would mean the timescale for promoter state transitions
is driven by topoisomerase kinetics, not by RNAP kinetics. From in vitro
measurements, the former are suggested to be of order minutes~\cite{Chong2014}.
Contrast this with model 2, where the state transitions are assumed to be
governed by RNAP, which, assuming a copy number per cell of order $10^3$, has a
diffusion-limited association rate $k_{on} \sim 10^2~\text{s}^{-1}$ to a target
promoter. Combined with known $K_d$'s of order $\mu$M, this gives an RNAP
dissociation rate $k_{off}$ of order $10^2$~s$^{-1}$. As we will show below,
however, there are some lingering puzzles with interpreting this supercoiling
hypothesis, so we leave it as a speculation and refrain from assigning definite
physical meaning to the two states in this model.
\mrm{I like this paragraph's philosphy, but I don't know if it should be cut out 
for the sake of shrinking the paper further.}

@@

To develop some intuition, consider model 4 in the limit $k^+ \ll k^- \lesssim
r$, which is roughly satisfied by the parameters inferred
in~\cite{Razo-Mejia2020}. In this limit, the system spends the majority of its
time in the inactive state, occasionally becoming active and making a burst of
transcripts. This should call to mind the well-known phenomenon of
transcriptional bursting, as observed in,
e.g.,~\cite{Golding2005,Chong2014,Sevier2016,Sepulveda2016}. Let us make this
correspondence more precise. The mean dwell time in the active state is $1/k^-$.
While in this state, transcripts are produced at a rate $r$ per unit time. So on
average, $r/k^-$ transcripts are produced before the system switches to the
inactive state. Once in the inactive state, the system dwells there for an
average time $1/k^+$ before returning to the active state and repeating the
process. $r/k^-$ resembles an average burst size, and $1/k^+$ resembles the time
interval between burst events. More precisely, $1/k^+$ is the mean time between
the end of one burst and the start of the next, whereas $1/k^+ + 1/k^-$ would be
the mean interval between the start of two successive burst events, but in the
limit $k^+ \ll k^-$, $1/k^+ + 1/k^- \approx 1/k^+$. Note that this limit ensures
that the waiting time between bursts is approximately exponentially distributed,
with mean set by the only timescale left in the problem, $1/k^+$. If instead it
were the case that $k^+ \sim k^-$, then the waiting time $1/k^+ + 1/k^-$ would
have a peaked distribution, but this does not appear to be the case for any of
datasets from~\cite{Jones2014}.

Let us now verify this intuition with a precise derivation to check that $r/k^-$
is in fact the mean burst size and to obtain the full burst size distribution.
Consider first a constant, known dwell time $T$ in the active state. Transcripts
are produced at a rate $r$ per unit time, so the number of transcripts $n$
produced during $T$ is Poisson distributed with mean $rT$, i.e.,
\begin{equation}
    P(n\mid T) = \frac{(rT)^n}{n!} \exp(-rT).
\end{equation}
Since the dwell time $T$ is unobservable from single molecule mRNA counts, we
actually want $P(n)$, the dwell time distribution with no conditioning on $T$.
Basic rules of probability theory tell us we can write $P(n)$ in terms of
$P(n\mid T)$ as
\begin{equation}
    P(n) =\int_0^\infty P(n\mid T) P(T) dT.
\end{equation}
But we know the dwell time distribution $P(T)$, which is exponentially
distributed according to
\begin{equation}
    P(T) = k^- \exp(-T k^-),
\end{equation}
so $P(n)$ can be written as
\begin{equation}
    P(n) = k^- \frac{r^n}{n!}
            \int_0^\infty T^n\exp[-(r + k^-)T]\,dT.
\end{equation}
A standard integral table shows $\int_0^\infty x^n e^{-ax}\,dx = n!/a^{n+1}$, so
\begin{equation}
    P(n) = \frac{k^- r^n}{(k^- + r)^{n+1}}
        = \frac{k^-}{k^- + r}
            \left(\frac{r}{k^- + r}\right)^n
        = \frac{k^-}{k^- + r}
            \left(1 - \frac{k^-}{k^- + r}\right)^n,
\end{equation}
which is exactly the geometric distribution with standard parameter
$\theta\equiv k^-/(k^- + r)$ and domain $n \in \{0, 1, 2, \dots\}$.
The mean of the geometric distribution, with this convention, is
\begin{align}
\langle n\rangle = \frac{1 - \theta}{\theta}
        = \left(1 - \frac{k^-}{k^- + r}\right)
                    \frac{k^- + r}{k^-}
        = \frac{r}{k^-},
\end{align}
exactly as we guessed intuitively above.

@@

Consider the time delay between the first and final RNAPs in a burst initiating
transcription (\textit{not} the time to complete transcripts, which potentially
could be much longer.) If this timescale is short compared to the typical search
timescale of transcription factors, then all is well. The estimates from
deHaseth et.\ al.~\cite{DeHaseth1998} put RNAP's diffusion-limited on rate
around $\sim\text{few}\times10^{-2}~\text{nM}^{-1}~\text{s}^{-1}$ and polymerase
loading as high as $1~\text{s}^{-1}$. Then for reasonable burst sizes of $<10$,
it is reasonable to guess that bursts might finish initiating on a timescale of
tens of seconds or less (with another 30-60 sec to finish elongation, but that
does not matter here). A transcription factor with typical copy number of order
10 (or less) would have a diffusion-limited association rate of order
$(10~\text{sec})^{-1}$ \cite{Hammar2014}. Higher copy number TFs tend to have
many binding sites over the genome, which should serve to pull them out of
circulation and keep their effective association rates from rising too large.
Therefore, there is \textit{perhaps} a timescale separation possible between
transcription factor association rates and burst durations, but this assumption
could very well break down, so we will have to keep it in mind when we infer
repressor rates from the Jones et.\ al.\ single-cell mRNA counts data
later~\cite{Jones2014}.
\mrm{Another great paragraph that could potentially be removed.}
